Community biomass and species richness temporal trends reveal contrasting assembly and disassembly processes in natural food-webs

### 2022-11-07 paper/appendix.Rmd
Cosmetic to pdf

```{r, echo=FALSE}
knitr::opts_chunk$set(
  cache = FALSE,
  collapse = TRUE,
  comment = "#>",
  #fig.dim = c(7, 7),
  fig.align = "center",
  fig.fullwidth = TRUE,
  fig.show = "hold",
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  results = TRUE
)
options(
  #kableExtra.latex.load_packages = FALSE,
  knitr.table.format = "latex"
)
```

>>>>

```{r, echo=FALSE}
knitr::opts_chunk$set(
  cache = FALSE,
  collapse = TRUE,
  comment = "#>",
  #fig.dim = c(7, 7),
  fig.align = "center",
  fig.fullwidth = TRUE,
  fig.show = "hold",
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  results = TRUE
)
options(
  #kableExtra.latex.load_packages = FALSE,
  knitr.table.format = "latex",
  knitr.graphics.rel_path = FALSE
)
```

### 2022-11-07 paper/manuscript.Rmd
Cosmetic to pdf

---
title: "Community biomass and species richness temporal trends reveal contrasting assembly and disassembly processes in natural food-webs"
author: Alain Danet, Loubna El Madouri, Maud Mouchet, Elisa Thébault and Colin Fontaine
date: \today
output:
#  bookdown::word_document2:
#    fontsize: 12pt
  bookdown::pdf_document2:
    fig_caption: true 
    keep_tex: true
    toc: true
    fontsize: 12pt
    header-includes:
       - \usepackage{natbib}
       - \usepackage{float}
       - \usepackage{multirow}
       - \usepackage{booktabs}
       - \usepackage{makecell}
       - \usepackage{setspace}
       - \doublespacing
       - \usepackage{lineno}
       - \linenumbers
    geometry: margin=2cm
    number_sections: true
bibliography: references.bib
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  cache = FALSE,
  collapse = TRUE,
  comment = "#>",
  #fig.dim = c(7, 7),
  fig.fullwidth = TRUE,
  fig.show = "hold",
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  results = TRUE,
  fig.path="figures/"
)
```

- Tentative titles:
  - How food-webs assembly and disassembly through time in natural settings?
    Lessons from freshwater fishes
  - Relationships among temporal trends of food-web structure, species richness
    and community biomass reveal contrasting assembly
    and disassembly processes
  - Temporal trends in food web structure according to richness and biomass
    reveal contrasting assembly and disassembly processes

@@

theme_set(mytheme)
options(knitr.table.format = "latex")
```

@@

There is rising recognition that biodiversity changes are
multidimensional[@mcgill_fifteen_2015; @blowes_geography_2019;
@hillebrand_biodiversity_2018; @blowes_local_nodate]. Studies found important
changes in species composition despite no changes in species richness in average
[@dornelas_assemblage_2014; @blowes_geography_2019], and highlighting the need
for further investigations on how biodiversity changes over time.

Yet species interactions are largely absent from biodiversity changes. Although
species interactions, described as ecological networks, drive for a large part
community composition, species coexistence, and ecosystem
functioning[@chesson_mechanisms_2000; @thebault_food-web_2003;
@oconnor_general_2017]. The structure of ecological networks can trigger or
impede perturbation propagations, i.e secondary extinctions[@paine_note_1969;
@dunne_network_2002; @srinivasan_response_2007;
@stouffer_compartmentalization_2011]. Higher connectance, which measures the
degree of generalism in a network, is predicted to increase robustness to
secondary extinctions[@dunne_network_2002; @sole_complexity_2001] and studies
found that higher trophic level was associated to a lower temporal variability
of community biomass in theoretical models and empirical
data[@shanafelt_stability_2018; @danet_species_2021]. Then, accounting for how
the structure of ecological networks changes over time according to other
biodiversity facets is crucial.

Th availability of ecological network timeseries is
scarce[@kortsch_disentangling_2021], such as predictions about how the structure
of ecological networks change according to changes in species richness come from
spatial comparison of networks. Such studies showed that connectance is
negatively related to species richness in small (i.e. below 20 species)
empirical food-webs [reviewed in @dunne_network_2006], which reflects
constraints on the number of realizable species interactions or an increasing
proportion of specialists in more rich foodwebs. Connectance has been shown to
scale positively or negatively with community biomass in theoretical
food-webs[@thebault_food-web_2003; @poisot_trophic_2013] while the average
trophic level was previously shown to scale negatively with species
richness[@danet_species_2021].

However, predicting how ecological networks will change in time according to
other biodiversity facets based on spatial comparison of ecological communities
is valid as long as assumption of the space-for-time substitution is valid as
well. Space-for-time substitution supposes a homology between variations in
space and in time, i.e. that spatial sequences correspond to temporal sequences.
This method have raised criticisms[@damgaard_critique_2019], with studies
showing analogous patterns in space and time[@blois_space_2013] or not
[@johnson_testing_2008; @adler_contrasting_2007]. Spatial turnover of species
composition, dispersal limitation, disturbance and stochastic processes have
been hypothesised to be responsible for the non-compliance of space-for-time
substitution[@adler_contrasting_2007; @blois_space_2013].

The increasing availability ecological temporal series and the development of
methods to reconstruct food-webs[@bonnaffe_comparison_2021;
@gravel_inferring_2013; @danet_species_2021] makes possible to overcome those
research gaps. We used a standardized monitoring of fish riverine communities on
the period 1995-2018 over France, documenting the individual size of fishes in
`r nb_st` sites during ten years or more. We reconstructed around `r nb_op`
food-webs based on allometric relationships and food diet data, and then
computed connectance, and average trophic level as well as community biomass and
species richness. We assessed how the local structure of food-webs changes over
time according to simultaneous changes in community biomass and species richness
over time and then if the relationships between changes in food-web structure,
community biomass and species richness, are the same over space and over time,
i.e. is the space-for-time substitution valid?

@@

```{r}
tar_load(glob_tps_trends_decade_no_drivers)
gbm <- get_effect_ci(resp = "log_bm_std", ci_lvl = "level:0.80", p = TRUE)
gr <- get_effect_ci(resp = "log_rich_std", ci_lvl = "level:0.80", p = TRUE)
gt <- get_effect_ci(resp = "w_trph_lvl_avg", ci_lvl = "level:0.80", p = FALSE)
gct <- get_effect_ci(resp = "ct_ff", ci_lvl = "level:0.80", p = FALSE)
```

@@

```{r sem, fig.cap = sem_cap}
knitr::include_graphics(here::here("figures", "fig2.png"))
```

@@

Using structural equation models to disentangle the direct and indirect
relationships among foodweb structure descriptors, we found a strong positive
relationship between the temporal trends of community biomass and of
species richness ($r_\delta = `r rdtps$bm$r`$, Fig. \@ref(fig:sem) a and b,
$r_\delta$: standardized slope coefficients, [95% CI]), meaning that both
temporal trends had the same direction in average. At the sites where species
richness decreased over time, community biomass also decreased over time. On the
other hand, at the sites where species richness increased over time, community
biomass also increased over time.

The temporal trends of connectance were positively linked with the ones of
community biomass ($r_\delta = `r rdtps$ct$bm`$), also meaning that both
temporal trends had the same direction in average. That is, the sites where
food-web connectance decreased over time also had community biomass that
decreased over time and reversely. It suggests that communities that decreased
in biomass over time lose highly connected trophic species over time, and
reversely that communities gaining biomass over time also gain highly connected
trophic species over time. We found a negative relationship between the temporal
trends of connectance and the ones of species richness
($r_\delta = `r rdtps$ct$r`$, Fig. \@ref(fig:sem)d). We further found the
species richness had an indirect positive effect on connectance, through
community biomass ($r_\delta = `r ridtps$ct$r`$). It resulted in a relatively
weak total negative effect of species richness on connectance
($r_\delta = `r rttps$ct$r`$), that we also found in random extinction
simulation (Fig. S2).

The temporal trends of average trophic level had a strong positive relationship
with the ones of community biomass ($r_\delta = `r rdtps$tlvl$bm`$, Fig.
\@ref(fig:sem)e), meaning that sites where community biomass decreased over
time, average trophic level decreased over time and reversely. It means that in
communities losing biomass over time, the average biomass of the food-web moved
down to lower trophic level and vice versa. However, we found that the temporal
trends of average trophic level had no direct relationship the ones of species
richness ($r_\delta = `r rdtps$tlvl$r`$, Fig. \@ref(fig:sem)f). It suggests that
in communities either losing or gaining species richness over time, the location
of average biomass along trophic level does not significantly change according
to species richness. However, the temporal trends of species richness had an
indirect positive relationship with the temporal trends of average trophic level
through community biomass ($r_\delta = `r ridtps$tlvl$r`$), resulting in a total
positive relationship ($r_\delta = `r rttps$tlvl$r`$).

@@

As for the temporal analysis, we found a positive relationship between
median community biomass and median species richness along the spatial gradient,
but half the strength as the temporal one ($r_\delta = `r rdsp$bm$r`$, Fig.
\@ref(fig:std)). We further found a positive relationship between median
connectance and median community biomass ($r_\delta = `r rdsp$ct$bm`$), as in
temporal trends analysis. However, we found a negative relationship between
median connectance and median species richness, but
three times stronger than in the temporal analysis ($r_\delta = `r rdsp$ct$r`$,
Fig. \@ref(fig:std)). We also found a positive indirect effect of median
richness on median connectance through median community biomass but a weak one
($r_\delta = `r ridsp$ct$r`$), resulting in a strong negative total effect of
median richness on median connectance ($r_\delta = `r rtsp$ct$r`$).

We found a positive relationship between median average trophic level and median
community biomass, but half the strength as in the temporal analysis ($r_\delta
= `r rdsp$tlvl$bm`$, Fig. \@ref(fig:std)). However, we found a strong negative
relationship between median average trophic level and median richness ($r_\delta
= `r rdsp$tlvl$r`$), in contrast with findings in temporal analysis. We also
found a positive indirect relationship between average trophic level and  median
richness through median community biomass (`r ridsp$tlvl$r`), resulting in a
strong total negative effect of median richness on median average trophic level
(`r rtsp$tlvl$r`).

@@

# Statistical analysis

We modelled the relationships among food-web structure, that is
connectance and average trophic level, and community biomass ($dB$) and species
richness ($dR$) with Structural Equation Models. We modelled a it such as species
richness is related to community biomass and food-web structure. In
turn community biomass was related to food-web structure. We computed one
structural equation model to assess the relationships among the temporal trends
and one for the relationships over the spatial gradient. As spatial variation of
food-web structure are partly driven by environmental gradients, so we
controlled relationships among variables by adding two PCA axis
synthesizing respectively (1) stream gradient from upstream to downstream and
(2) enrichment (Fig. \@ref(fig:matmet)). Enrichment was characterized by the
average Biological Oxygen Demand [BOD, @mallin_factors_2006][^2]. We further
added a random effect of the hydrographic basin on the intercept.

@@

All the statistical models displayed low multicollinearity (Variance inflation
factor < 3, Table SXX).  We visually inspected the residuals of the linear
models.

@@

@@

# Extinction simulation in the metaweb

>>>>

---
title: "Community biomass and species richness temporal trends reveal contrasting assembly and disassembly processes in natural food-webs"
author: Alain Danet, Loubna El Madouri, Maud Mouchet, Elisa Thébault and Colin Fontaine
date: \today
output:
#  bookdown::word_document2:
#    fontsize: 12pt
  bookdown::pdf_document2:
    fig_caption: true 
    keep_tex: true
    toc: false 
    fontsize: 12pt
header-includes:
   - \usepackage{natbib}
   - \usepackage{float}
   - \usepackage{multirow}
   - \usepackage{booktabs}
   - \usepackage{makecell}
   - \usepackage{setspace}
   - \doublespacing
   - \usepackage{lineno}
   - \linenumbers
geometry: margin=2cm
number_sections: true
bibliography: references.bib
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  cache = FALSE,
  collapse = TRUE,
  comment = "#>",
  #fig.dim = c(7, 7),
  fig.fullwidth = TRUE,
  fig.show = "hold",
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  results = TRUE,
  fig.path="figures/"
)
options(
  knitr.table.format = "latex",
  knitr.graphics.rel_path = FALSE
)
```

@@

theme_set(mytheme)
```

@@

There is rising recognition that biodiversity changes are
multidimensional [@mcgill_fifteen_2015; @blowes_geography_2019;
@hillebrand_biodiversity_2018; @blowes_local_nodate]. Studies found important
changes in species composition despite no changes in species richness in average
[@dornelas_assemblage_2014; @blowes_geography_2019], and highlighting the need
for further investigations on how biodiversity changes over time.

Yet species interactions are largely absent from biodiversity changes. Although
species interactions, described as ecological networks, drive for a large part
community composition, species coexistence, and ecosystem
functioning [@chesson_mechanisms_2000; @thebault_food-web_2003;
@oconnor_general_2017]. The structure of ecological networks can trigger or
impede perturbation propagations, i.e secondary extinctions [@paine_note_1969;
@dunne_network_2002; @srinivasan_response_2007;
@stouffer_compartmentalization_2011]. Higher connectance, which measures the
degree of generalism in a network, is predicted to increase robustness to
secondary extinctions [@dunne_network_2002; @sole_complexity_2001] and studies
found that higher trophic level was associated to a lower temporal variability
of community biomass in theoretical models and empirical
data [@shanafelt_stability_2018; @danet_species_2021]. Then, accounting for how
the structure of ecological networks changes over time according to other
biodiversity facets is crucial.

Th availability of ecological network timeseries is
scarce [@kortsch_disentangling_2021], such as predictions about how the structure
of ecological networks change according to changes in species richness come from
spatial comparison of networks. Such studies showed that connectance is
negatively related to species richness in small (i.e. below 20 species)
empirical food-webs [reviewed in @dunne_network_2006], which reflects
constraints on the number of realizable species interactions or an increasing
proportion of specialists in more rich foodwebs. Connectance has been shown to
scale positively or negatively with community biomass in theoretical
food-webs [@thebault_food-web_2003; @poisot_trophic_2013] while the average
trophic level was previously shown to scale negatively with species
richness [@danet_species_2021].

However, predicting how ecological networks will change in time according to
other biodiversity facets based on spatial comparison of ecological communities
is valid as long as assumption of the space-for-time substitution is valid as
well. Space-for-time substitution supposes a homology between variations in
space and in time, i.e. that spatial sequences correspond to temporal sequences.
This method have raised criticisms [@damgaard_critique_2019], with studies
showing analogous patterns in space and time [@blois_space_2013] or not
[@johnson_testing_2008; @adler_contrasting_2007]. Spatial turnover of species
composition, dispersal limitation, disturbance and stochastic processes have
been hypothesised to be responsible for the non-compliance of space-for-time
substitution [@adler_contrasting_2007; @blois_space_2013].

The increasing availability ecological temporal series and the development of
methods to reconstruct food-webs [@bonnaffe_comparison_2021;
@gravel_inferring_2013; @danet_species_2021] makes possible to overcome those
research gaps. We used a standardized monitoring of fish riverine communities on
the period 1995-2018 over France, documenting the individual size of fishes in
`r nb_st` sites during ten years or more. We reconstructed around `r nb_op`
food-webs based on allometric relationships and food diet data, and then
computed connectance, and average trophic level as well as community biomass and
species richness. We assessed how the local structure of food-webs changes over
time according to simultaneous changes in community biomass and species richness
over time and then if the relationships between changes in food-web structure,
community biomass and species richness, are the same over space and over time,
i.e. is the space-for-time substitution valid?

@@

```{r, results=FALSE}
tar_load(glob_tps_trends_decade_no_drivers)
glob_tps_trends_decade_no_drivers
gbm <- get_effect_ci(resp = "log_bm_std", ci_lvl = "level:0.80", p = TRUE)
gr <- get_effect_ci(resp = "log_rich_std", ci_lvl = "level:0.80", p = TRUE)
gt <- get_effect_ci(resp = "w_trph_lvl_avg", ci_lvl = "level:0.80", p = FALSE)
gct <- get_effect_ci(resp = "ct_ff", ci_lvl = "level:0.80", p = FALSE)
```

@@

```{r sem, fig.cap = sem_cap}
knitr::include_graphics(here::here("figures", "fig2.pdf"))
```

@@

Using structural equation models to disentangle the direct and indirect
relationships among foodweb structure descriptors, we found a strong positive
relationship between the temporal trends of community biomass and of
species richness ($r_\delta =$ `r rdtps$bm$r`, Fig. \@ref(fig:sem) a and b,
$r_\delta$: standardized slope coefficients, [95% CI]), meaning that both
temporal trends had the same direction in average. At the sites where species
richness decreased over time, community biomass also decreased over time. On the
other hand, at the sites where species richness increased over time, community
biomass also increased over time.

The temporal trends of connectance were positively linked with the ones of
community biomass ($r_\delta =$ `r rdtps$ct$bm`), also meaning that both
temporal trends had the same direction in average. That is, the sites where
food-web connectance decreased over time also had community biomass that
decreased over time and reversely. It suggests that communities that decreased
in biomass over time lose highly connected trophic species over time, and
reversely that communities gaining biomass over time also gain highly connected
trophic species over time. We found a negative relationship between the temporal
trends of connectance and the ones of species richness
($r_\delta =$ `r rdtps$ct$r`, Fig. \@ref(fig:sem)d). We further found the
species richness had an indirect positive effect on connectance, through
community biomass ($r_\delta =$ `r ridtps$ct$r`). It resulted in a relatively
weak total negative effect of species richness on connectance
($r_\delta =$ `r rttps$ct$r`), that we also found in random extinction
simulation (Fig. S2).

The temporal trends of average trophic level had a strong positive relationship
with the ones of community biomass ($r_\delta =$ `r rdtps$tlvl$bm`, Fig.
\@ref(fig:sem)e), meaning that sites where community biomass decreased over
time, average trophic level decreased over time and reversely. It means that in
communities losing biomass over time, the average biomass of the food-web moved
down to lower trophic level and vice versa. However, we found that the temporal
trends of average trophic level had no direct relationship the ones of species
richness ($r_\delta =$ `r rdtps$tlvl$r`, Fig. \@ref(fig:sem)f). It suggests that
in communities either losing or gaining species richness over time, the location
of average biomass along trophic level does not significantly change according
to species richness. However, the temporal trends of species richness had an
indirect positive relationship with the temporal trends of average trophic level
through community biomass ($r_\delta =$ `r ridtps$tlvl$r`), resulting in a total
positive relationship ($r_\delta =$ `r rttps$tlvl$r`).

@@

As for the temporal analysis, we found a positive relationship between
median community biomass and median species richness along the spatial gradient,
but half the strength as the temporal one ($r_\delta =$ `r rdsp$bm$r`, Fig.
\@ref(fig:std)). We further found a positive relationship between median
connectance and median community biomass ($r_\delta =$ `r rdsp$ct$bm`), as in
temporal trends analysis. However, we found a negative relationship between
median connectance and median species richness, but
three times stronger than in the temporal analysis ($r_\delta =$ `r rdsp$ct$r`,
Fig. \@ref(fig:std)). We also found a positive indirect effect of median
richness on median connectance through median community biomass but a weak one
($r_\delta =$ `r ridsp$ct$r`), resulting in a strong negative total effect of
median richness on median connectance ($r_\delta =$ `r rtsp$ct$r`).

We found a positive relationship between median average trophic level and median
community biomass, but half the strength as in the temporal analysis ($r_\delta
=$ `r rdsp$tlvl$bm`, Fig. \@ref(fig:std)). However, we found a strong negative
relationship between median average trophic level and median richness ($r_\delta
=$ `r rdsp$tlvl$r`), in contrast with findings in temporal analysis. We also
found a positive indirect relationship between average trophic level and  median
richness through median community biomass (`r ridsp$tlvl$r`), resulting in a
strong total negative effect of median richness on median average trophic level
(`r rtsp$tlvl$r`).

@@

## Statistical analysis

We modelled the relationships among food-web structure, that is
connectance and average trophic level, and community biomass ($dB$) and species
richness ($dR$) with Structural Equation Models. We modelled a it such as species
richness is related to community biomass and food-web structure. In
turn community biomass was related to food-web structure. We computed one
structural equation model to assess the relationships among the temporal trends
and one for the relationships over the spatial gradient. As spatial variation of
food-web structure are partly driven by environmental gradients, so we
controlled relationships among variables by adding two PCA axis
synthesizing respectively (1) stream gradient from upstream to downstream and
(2) enrichment (Fig. \@ref(fig:matmet)). Enrichment was characterized by the
average Biological Oxygen Demand [BOD, @mallin_factors_2006]. We further
added a random effect of the hydrographic basin on the intercept.

@@

All the statistical models displayed low multicollinearity (Variance inflation
factors < 3, Table SXX).  We visually inspected the residuals of the linear
models.

@@

## Extinction simulation in the metaweb

@@
